Variational data assimilation for the wave equation in heterogeneous media

### 2025-12-04 review/response/letter.tex

First of all, thank you for considering our article "Variational data assimilation for the wave equation in heterogeneous media: Numerical investigation of stability" for publication in \textit{Communications on Applied Mathematics and Computation}. We also thank the reviewers for their valuable comments and suggestions for improvement. Below, we address their points one after another. \\

%------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------

%------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------
\subsection*{Referee 1}	
%------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------

\begin{enumerate}
  \item The Geometric Control Condition (GCC) was originally introduced in [8]. However, the
  manuscript does not clarify how GCC satisfaction is verified in the specific numerical
  simulations. For completeness, I recommend including a precise definition of the GCC.
  \reply{\textbf{add a bit of detail, lightlike..., precise definition to technical...} @Janosch}
  \item On page 7, could the authors clarify whether $\bm{Z}_h$ belongs to the space $\mathcal{W}_{h, \Delta t}^{k^{\ast},q^{\ast}}$?
  \reply{Indeed, it does belong to $\mathcal{W}_{h, \Delta t}^{k^{\ast},q^{\ast}}$. We have added this after the definition of the Lagrangian $\mathcal{L}$.}
  \item In the numerical experiments, what values are chosen for the parameters $k, q, k^{\ast}$ and
  $q^{\ast}$? The current description appears to only specify the value of $k$.
  \reply{We have added the values of $q, k^{\ast}, q^{\ast}$ for the numerical experiments. \textbf{@ Janosch, why do we choose them that way?}}
  \item How is the “best approximation” determined in the numerical experiments? Please explain the criteria or method used to obtain it.
  \reply{In the numerical experiments, we compare the approximation error of our method with the $L^2$-best approximation error, i.e. we calculate the $L^2$-orthogonal projection of the exact solution onto the discrete space. We have added a footnote.}
  \item On page 18, the black line in Figure 5, does it represent the exact solution at time 0.5
  or at time 1? It seems to depict the solution at $T = 1.0$, where the approximation
  matches the exact solution well. The exact solution at $T = 0.5$ does not appear to be shown.
  \reply{We compute the approximation twice, once with final time $T = 0.5$ and once with final time $T = 1.0$, but compare it for both cases at $t = 0.5$. The figure illustrates that the approximation is not reliable if the final time is not large enough to satisfy the GCC. We have slightly adjusted the text and the figure to avoid any misunderstandings.}
  \item In the two-dimensional case, only a square data domain is considered. A more interesting scenario would involve a curved interface, such as a circular interface. Could the
  authors provide related numerical results for such a case?
\end{enumerate}

\subsection*{Referee {2}}	
%------------------------------------------------------------------------------------------------
%------------------------------------------------------------------------------------------------
\begin{enumerate}
  \item How did you numerically verify Assumption 1? Specifically, how were the two
  inequalities in Assumption 1 verified?
  \reply{We proved that the assumptions gives us convergence rates and check the convergence rates numerically..., refer to Remark 2, numerical method gives stability of the continuous problem}
  \item What is the relationship between the GCC and condition (1.3)? In the numerical tests in Section 4.1, you use condition (1.3) to imply the GCC. It appears that you have verified stability under a much stronger condition. I am wondering what would happen if you directly let the data satisfy the GCC, as you did in Section 4.2. Moreover, could you please explain why $T = 0.75$ was chosen in
  Section 4.2? Would a smaller $T$ work?
  \reply{In 1D: (1.3) is equivalent to the GCC... \textbf{2D not sure, think about this after modifications from Ref 1}}
  \item Could you please explain why you adopted the discontinuous-in-time FEM
  method? What are the advantages of this method for the problem under consideration?
  \reply{@ Janosch}
  \item Please clarify the choices of $q$, $k^{\ast}$ , and $q^{\ast}$ in the numerical experiments.
  \reply{We have added the values of $q, k^{\ast}, q^{\ast}$ for the numerical experiments.}
  \item Please clarify the conditions of Lemma 3.1. Does it require the GCC or condition
  (1.3)?
  \reply{It requires (1.3), which is required for unique solvability anyway. We have added a short sentence in the beginning of Sec. 3.1 that we assume (1.3) to be satisfied.}
  \item The authors are encouraged to provide a detailed proof of Lemma 3.3.
  \reply{We added a proof of Lemma 3.3.} 
  \item What would happen if $c_1$ becomes larger?
  \reply{becomes harder to solve, see conclusion and discussion in 4.2...}
  \item Please clarify the meaning of “the $L^2$ -best approximation errors” in the numerical section.
  \reply{We have added a footnote (see above, point 4.).}
  \item Typo: last line on page 19, “$L^2$ -bestapproximation” $\to$ “$L^2$ -best approximation”.
  \reply{We thank the review for spotting this typo. We have corrected it.}
\end{enumerate}

### 2025-12-07 tex/main.tex

\subsection{Examples in one space dimension}
 We consider the unit interval $\Omega \coloneqq (0,1) \subset \mathbb{R}$ partitioned into $2^{L+1}$, $L \in \mathbb{N}$, uniform elements of size $h = 1/2^{L+1}$. Unless specified otherwise, we choose the time step as $\Delta t = h$.

@@

In \Cref{fig:jumpCoefs:contrast:2.5}, we observe that the errors $\Vert u-L_{\Delta t} \ul_1 \Vert_{L^\infty(0,T;L^2(\Omega_R))}$ and $\Vert \partial_t (u-L_{\Delta t} \ul_1) \Vert_{L^2(0,T;L^2(\Omega_R))}$ converge with optimal order $\mathcal{O}(h^k)$ for $T = 0.5 > 0.35$. However, if $T = 0.1 < 0.35$, the order of convergence degrades, while the respective best approximation errors still converge with optimal order. This suggests that the GCC seems to imply Lipschitz stability even in the heterogeneous case, and therefore \cref{assum:LipschitzStability} is indeed justified.

@@

As a second example, we consider the case where the data is only given in one half of the interval, i.e. $\omega = [0,0.3]$. The requirement \eqref{eq:Tcondition} and thus the GCC is fulfilled if 
\begin{equation}
    T > 2 \left[ (p_2 - p_1) + \frac{p_1 - 0.3 + 1 - p_2}{c_1} \right]. 
\end{equation}
We consider $c_1 \in \{ 2.5, 7.5\}$, for which the requirement $T > \{ 0.89,0.65 \}$ ensures the GCC. \Cref{fig:jumpCoefs:multipleJumps:noGCC} compares the results for polynomial degree $k = 3$ on the refinement level $L = 4$ with a fixed time step $\Delta t = 1/32$, for two final times $T = 0.5$ and $T = 1.0$. For both contrasts, the approximate solution matches the exact solution well if the GCC is fulfilled ($T = 1.0$), but shows significant deviations when it is not ($T = 0.5$). 
The deviation is more extreme for contrast $c_1 = 2.5$, where the gap between the considered time $T = 0.5$ and the threshold $T > 0.89$ is larger.

@@

            \end{groupplot}
        \end{tikzpicture}
         \pgfplotslegendfromname{named}
    \end{center}
    \caption{For fixed time step $\Delta t = 1/32$, refinement level $L = 4$, and polynomial degree $k = 3$, we compare the exact solution \eqref{eq:1D:exact:simpleMult} and approximated solution computed with $T = 1.0$ and $T = 0.5$ at time $t = 0.5$. We set $c_2 = 1.0$ and consider the cases $c_1 = c_3 = 2.5$, $n = 1$ (left) and $c_1 = c_3 = 7.5$, $n = 3$ (right).} 
    \label{fig:jumpCoefs:multipleJumps:noGCC}
  \end{figure}

@@

We define the data domain $\omega \coloneqq \Omega \setminus [0.25,0.75]^2$ and note that $\omega$ fulfills the GCC. To ensure that the data domain and the interface $\Gamma$ are meshed exactly, we consider a quadrilateral mesh with $2^{L+1}$, $L \in \{1,2,3,4\}$, elements in each direction. We consider the final time $T = 0.75$  and set the time step size to $\Delta t = (1/2)^{L+1}$. As before, we fix the wave speed $c_2 = 1$ and consider different values for $c_1$ to control the contrast.

The results of the experiment are presented in \cref{fig:2DjumpCoefs}. 
With contrast $c_1 = 2.5$, we observe quasi-optimal convergence with order  $\mathcal{O}(h^k)$ for polynomial degrees $k \in \{2,3\}$ for the errors $\Vert \dt(u-L_{\Delta t} \ul_1) \Vert_{L^2(0,T;L^2(\Omega))}$ and $\Vert u-L_{\Delta t} \ul_1 \Vert_{L^\infty(0,T;L^2(\Omega))}$. 
Similar as already observed in the one-dimensional case in \cref{fig:jumpCoefs:increasingcontrast}, we observe that these errors are more sensitive to increasing contrast $c_1$ than the corresponding best approximation errors. 
In particular, the ratio for $\Vert u-L_{\Delta t} \ul_1 \Vert_{L^\infty(0,T;L^2(\Omega))}$ with $k=2$ appears to grow as $\mathcal{O}(c_1^{3/2})$, which renders it very demanding to maintain a reasonable accuracy as the contrast increases.

>>>>

\subsection{Examples in one space dimension}
 We consider the unit interval $\Omega \coloneqq (0,1) \subset \mathbb{R}$ partitioned into $2^{L+1}$, $L \in \mathbb{N}$, uniform elements of size $h = 1/2^{L+1}$. Unless specified otherwise, we choose the time step as $\Delta t = h$. \changed{Further, we set $q = k$ and choose $k^{\ast} = 1$, $q^{\ast} = 1$ if $k = 1$ and $q^{\ast} = 0$ if $k > 1$.}

@@

In \Cref{fig:jumpCoefs:contrast:2.5}, we observe that the errors $\Vert u-L_{\Delta t} \ul_1 \Vert_{L^\infty(0,T;L^2(\Omega_R))}$ and $\Vert \partial_t (u-L_{\Delta t} \ul_1) \Vert_{L^2(0,T;L^2(\Omega_R))}$ converge with optimal order $\mathcal{O}(h^k)$ for $T = 0.5 > 0.35$. However, if $T = 0.1 < 0.35$, the order of convergence degrades, while the respective $L^2$-best approximation\textsuperscript{3}\footnotemark \footnotetext{\textsuperscript{3}\changed{We compare with the $L^2$-best approximation error, where we solve for the $L^2$-orthogonal projection of the exact solution onto the discrete space, as a measure of the approximation quality of the discrete space.}} errors still converge with optimal order. This suggests that the GCC seems to imply Lipschitz stability even in the heterogeneous case, and therefore \cref{assum:LipschitzStability} is indeed justified.

@@

As a second example, we consider the case where the data is only given in one half of the interval, i.e. $\omega = [0,0.3]$. The requirement \eqref{eq:Tcondition} and thus the GCC is fulfilled if 
\begin{equation}
    T > 2 \left[ (p_2 - p_1) + \frac{p_1 - 0.3 + 1 - p_2}{c_1} \right]. 
\end{equation}
We consider $c_1 \in \{ 2.5, 7.5\}$, for which the requirement $T > \{ 0.89,0.65 \}$ ensures the GCC. \Cref{fig:jumpCoefs:multipleJumps:noGCC} compares the results for polynomial degree $k = 3$ on the refinement level $L = 4$ with a fixed time step $\Delta t = 1/32$, for two final times $T = 0.5$ and $T = 1.0$. \changed{We compare the approximation of the exact solution for both final times at time $t = 0.5$}.
For both contrasts, the approximate solution matches the exact solution well if the GCC is fulfilled ($T = 1.0$), but shows significant deviations when it is not ($T = 0.5$). 
The deviation is more extreme for contrast $c_1 = 2.5$, where the gap between the considered time $T = 0.5$ and the threshold $T > 0.89$ is larger.

@@

            \end{groupplot}
        \end{tikzpicture}
         \pgfplotslegendfromname{named}
    \end{center}
    \caption{For fixed time step $\Delta t = 1/32$, refinement level $L = 4$, and polynomial degree $k = 3$, we compare the exact solution \eqref{eq:1D:exact:simpleMult} and approximated solution computed with \changed{final time} $T = 1.0$ and $T = 0.5$ at time $t = 0.5$. We set $c_2 = 1.0$ and consider the cases $c_1 = c_3 = 2.5$, $n = 1$ (left) and $c_1 = c_3 = 7.5$, $n = 3$ (right).} 
    \label{fig:jumpCoefs:multipleJumps:noGCC}
  \end{figure}

@@

We define the data domain $\omega \coloneqq \Omega \setminus [0.25,0.75]^2$ and note that $\omega$ fulfills the GCC. To ensure that the data domain and the interface $\Gamma$ are meshed exactly, we consider a quadrilateral mesh with $2^{L+1}$, $L \in \{1,2,3,4\}$, elements in each direction. We consider the final time $T = 0.75$  and set the time step size to $\Delta t = (1/2)^{L+1}$. As before, we fix the wave speed $c_2 = 1$ and consider different values for $c_1$ to control the contrast. \changed{Further, we choose the parameters $q = k^{\ast} = q^{\ast} = k$.}

The results of the experiment are presented in \cref{fig:2DjumpCoefs}. 
With contrast $c_1 = 2.5$, we observe quasi-optimal convergence with order  $\mathcal{O}(h^k)$ for polynomial degrees $k \in \{2,3\}$ for the errors $\Vert \dt(u-L_{\Delta t} \ul_1) \Vert_{L^2(0,T;L^2(\Omega))}$ and $\Vert u-L_{\Delta t} \ul_1 \Vert_{L^\infty(0,T;L^2(\Omega))}$. 
Similar as already observed in the one-dimensional case in \cref{fig:jumpCoefs:increasingcontrast}, we observe that these errors are more sensitive to increasing contrast $c_1$ than the corresponding $L^2$-best approximation errors. 
In particular, the ratio for $\Vert u-L_{\Delta t} \ul_1 \Vert_{L^\infty(0,T;L^2(\Omega))}$ with $k=2$ appears to grow as $\mathcal{O}(c_1^{3/2})$, which renders it very demanding to maintain a reasonable accuracy as the contrast increases.
